Equivalence Hypergraphs: E-Graphs for Monoidal Theories

### 2024-12-24 LICS/sections/appendix.tex
makefile added

\begin{remark}
    The first assumption in~\ref{pushout:assumptions} can be dropped with extra requirements that for all egdes $e_{i}, e_{j} \in E_{Z}$:
    \begin{itemize}
        \item $[f_{E}(e_{i})) = [f_{E}(e_{j}))$ and $[g_{E}(e_{i})) = [g_{E}(e_{j}))$ \textcolor{red}{This has to be dropped}
        \item \textcolor{red}{Find appropriate conditions}
    \end{itemize}
\end{remark}

\begin{remark}
    If in a span $Y \xleftarrow{g} Z \xrightarrow{f} X$ of morphisms in $\catname{EHyp}(\Sigma)$ the morphisms $f$ and $g$ are monos, then the arrows $f'$ and $g'$ of the pushout $Y \xrightarrow{g'} X +_{f,g} Y \xleftarrow{f'} X$ are also monos.
\end{remark}
This is because $[]_{V}$ identifies two vertices $v_1$ and $v_2$ only if there exists a vertex $z$ in $V_{Z}$ such that $f_{V};\iota_{1,V}(z) = v_1$ and $g_{V};\iota_{2,V}(z) = v_2$ and therefore $f' = \iota_{1};[]$ and $g' = \iota_{2};[]$ are monos.

@@

    where $V$s and $E$s are sets of vertices and hyperedges respectively, marked squares are pushouts and $\mathcal{L}^{\bot}$s are pushout complements.
    Because $i+j$ is discrete, $E_{\mathcal{G}}$ is the disjoint union of $E_{\mathcal{L}}$ and $E_{\mathcal{L}^{\bot}}$ and hence $E_{\mathcal{L}^{\bot}} = E_{\mathcal{G}} \setminus E_{\mathcal{L}}$.
    Then, since $m$ and $[c_1,c_2]$ are monos, the first square can be rewritten as follows.

@@

    \textcolor{red}{The same reasoning applies to show $E_{\mathcal{L}^{\bot}}$} is unique.
    One can further observe that $g_{V} = h_{V,ext};h_{V,int} + id_{n}$ and that $h_{V,ext}$ is mono and $h_{V,int} = [h_1,h_2]$ where $h_1,h_2$ are mono.
    The above shows that edges and vertices in the pushout complement are uniquely determined.

@@

    Because $g$ is a homomorphism, it must be the case that $g_{V}^{*}(s_{\mathcal{L^{\bot}}}(e)) = s_{\mathcal{G}}(g_{E}(e))$ and $s_{\mathcal{L^{\bot}}}(e) \subseteq (g_{V}^{-1})^{*}(s_{\mathcal{G}}(g_{E}(e)))$.
    Since $g_{V} = h_{V,ext};[h_1,h_2] + id_{n}$ the inverse image of $g_{V}$ contains at most two elements: one with the pre-image in $i$ and another with the pre-image in $j$.
    If $g_{V}$ identified more than two vertices it would mean that $[h_1,h_2]$ identified more than two vertices which would mean that two of them come both either from $i$ or $j$ which would imply that either $h_1$ or $h_2$ is not mono.
    If the inverse image contains exactly one vertex then the $s_{\mathcal{L^{\bot}}}(e)$ is uniquely fixed, otherwise suppose it contains $\{v_1,v_2\}$ such that $v_1 \in i$ and $v_2 \in j$.
    However, $v_1$ can not be in $s_{\mathcal{L^{\bot}}}(e)$ as per the (6) and (7) conditions of the boundary complement~\ref{def:boundary_new} the image of $j$ in $\mathcal{L}^{\bot}$ should consist of vertices of in-degree 0.

    Similarly for $t_{\mathcal{L}^{\bot}}$.

@@

    which implies $<_{2}(\iota_{V_{\mathcal{L}^{\bot}}}(v)) = <_{1}(\iota_{V_{\mathcal{L}^{\bot}}}(v))$ for all $v$ such that $[v) \not = \varnothing$ since $g_{E}$ is mono.

@@

    Clearly there is a morphism from $\mathcal{L}^{\bot}$ to $\mathcal{G}'$, i.e. the morphism $g' = (g_{V},g_{E})$ that has the same action on edges and vertices as the morphism $g$ does, because $v'$ such that $[v') \not = \varnothing$ can be mapped to $g_{V}(v')$ such that $[g_{V}(v')) \not = \varnothing$.

@@

    Finally, let's suppose there exist
    \[
        \mathcal{L}^{\bot}_{1} = (V_{\mathcal{L}^{\bot}}, E_{\mathcal{L}^{\bot}}, s,t, <, \consistency_{1})
    \quad
    \text{and}
    \quad
        \mathcal{L}^{\bot}_{2} = (V_{\mathcal{L}^{\bot}},V_{\mathcal{L}^{\bot}},s,t,<, \consistency_{2})
    \]
    Because $g$ is a homomorphism, we have
    \[
        [~g_{V};\iota_{V_{\mathcal{G}}}, g_{E};\iota_{E_{\mathcal{G}}}~]^{*}(\consistency_{\mathcal{L}^{\bot}_{1}}(\iota_{V_{\mathcal{L}^{\bot}}}(v))) \subseteq \consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v))
    \]
    \[
        [~g_{V};\iota_{V_{G}}, g_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{L}^{\bot}_{2}}(\iota_{V_{\mathcal{L}^{\bot}}}(v))) \subseteq \consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v))
    \]
    \begin{itemize}
        \item If $\consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v)) = \varnothing$ then both left-hand sides should be $\varnothing$ and therefore the relations are equal.
        \item Suppose that $\consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v)) \not = \varnothing$ which further means that $[g_{V};\iota_{V_{\mathcal{G}}}(v)) \not = \varnothing$.
              $\consistency_{1} \not = \consistency_{2}$ implies there exist $v, v' \in \mathcal{L}^{\bot}$ such that $v \consistency_{1} v'$ and $v \not \consistency_{2} v'$ for $[v) = [v') \not = \varnothing$.
              Then there are connected components $C_1, C_2$ in $\mathcal{L}_{2}^{\bot}$ such that for all $x \in C_1$ and $y \in C_2$ $x \not \consistency_{2} y$ and $v \in C_1$, $v' \in C_2$.
              Consider $g_2 = (g_{V},g_{E}) : \mathcal{L}_2^{\bot} \to \mathcal{G}$ and construct $\mathcal{G'}$ that is obtained from $\mathcal{G}$ by making $g(x) \not \consistency g(y)$ for $x \in C_1$ and $y \in C_2$.
              There is a morphism $g_2' = g_2 : \mathcal{L}_2^{\bot} \to \mathcal{G}'$. Like above, the only case when such morphism may not exist is when for $v' \not \consistency v$ $g_{V}(v) = g_{V}(v')$.
              This implies $v, v'$ have a pre-image in $i + j$ and by definition it must be $v' \consistency v$.
              Similarly, there is a morphism $m' : \mathcal{L} \to \mathcal{G}'$. The only case when it may not exist is when for $v_1, v_1' \in \mathcal{L}$ such that $v_1 \consistency v_1'$, $m_{V}(v_1) = g_{V}(v_2)$ and $m_{V}(v_1') = g_{V}(v_2')$ where $v_2 \in C_1$ and $v_2' \in C_2$.
              It implies $v_1, v_1'$ have pre-image in $i + j$ and by definition of a boundary-complement the image of $i + j$ should consist of top-level vertices exclusively.
              This results in the diagram below there is no morphism $u : G \to G'$, as the existence of such a morphism implies $u(g(x)) = u(g'(x))$, $u(g(y)) = u(g'(y))$ for $x \in C_1$ and $y \in C_2$ and $u(g(x)) \consistency u(g(y))$ and $u(g'(x)) \not \consistency u(g'(y))$.
              This contradicts the fact that $\mathcal{G}$ is a pushout.
              Hence, $\consistency_{1} = \consistency_{2}$ and $\mathcal{L}_{2}^{\bot} = \mathcal{L}_{1}^{\bot}$.
    \end{itemize}
    Ultimately, we have shown that $\mathcal{L}^{\bot}$ is unique up to isomorphism.
\end{proof}

@@

>>>>

\begin{remark}
    The first assumption in~\ref{pushout:assumptions} can be weakened by allowing $E_{Z}$ to contain unlabelled edges with no inputs and outputs such that:
    \begin{itemize}
        \item $[f_{E}(e_{i})) = [f_{E}(e_{j}))$ and $[g_{E}(e_{i})) = [g_{E}(e_{j}))$ for all $e_{i},e_{j} \in E_{Z}$ 
        \item If $[f_{E}(e)) \not = \varnothing$ then $[g_{E}(e)) = \varnothing$ and if $[g_{E}(e)) \not = \varnothing$ then $[f_{E}(e)) = \varnothing$.
        \item $\consistency(f_{E}(e_{i})) = \consistency(f_{E}(e_{j}))$ and $\consistency(g_{E}(e_i)) = \consistency(g_{E}(e_j))$ for all $e_i,e_j$ in $E_{Z}$.
    \end{itemize}
\end{remark}

\begin{remark}
    If in a span $Y \xleftarrow{g} Z \xrightarrow{f} X$ of morphisms in $\catname{EHyp}(\Sigma)$ the morphisms $f$ and $g$ are monos, then the arrows $f'$ and $g'$ of the pushout $Y \xrightarrow{g'} X +_{f,g} Y \xleftarrow{f'} X$ are also monos.
\end{remark}
This is because $[]_{V}$ identifies two vertices $v_1$ and $v_2$ only if there exists a vertex $z$ in $V_{Z}$ such that $f_{V};\iota_{1,V}(z) = v_1$ and $g_{V};\iota_{2,V}(z) = v_2$ and therefore $f' = \iota_{1};[]$ and $g' = \iota_{2};[]$ are monos.
Similarly for edges.

@@

    where $V$s and $E$s are sets of vertices and hyperedges respectively, marked squares are pushouts and $\mathcal{L}^{\bot}$s are pushout complements and where $E_{0}$ is either empty or contains edges with no inputs and outputs only.
    % Because $i+j$ is discrete, $E_{\mathcal{G}}$ is the disjoint union of $E_{\mathcal{L}}$ and $E_{\mathcal{L}^{\bot}}$ and hence $E_{\mathcal{L}^{\bot}} = E_{\mathcal{G}} \setminus E_{\mathcal{L}}$.
    Since $m$ and $[c_1,c_2]$ are monos, the first square can be rewritten as follows.

@@

    \textcolor{red}{The same reasoning applies to show that $E_{\mathcal{L}^{\bot}}$} is unique and hence the sets of edges and vertices are uniquely determined in the pushout complement.
    One can observe that $g_{V} = h_{V,ext};h_{V,int} + id_{n}$ and that $h_{V,ext}$ is mono and $h_{V,int} = [h_1,h_2]$ where $h_1,h_2$ are mono.

@@

    Suppose, on contrary, that there exist
    \mathcal{L}_{1}^{\bot} = \langle V_{\mathcal{L}^{\bot}}, E_{\mathcal{L}^{\bot}}, s_{1} \rangle \qquad \mathcal{L}_{2}^{\bot} = \langle V_{\mathcal{L}^{\bot}}, E_{\mathcal{L}^{\bot}}, s_{2} \rangle

    such that there is $e \in E_{\mathcal{L}^{\bot}}$ and $s_{1}(e) \not = s_{2}(e)$.
    As $g$ is a homomorphism, it must be the case that $g^{*}_{V}(s_{1}(e)) = g^{*}_{V}(s_{2}(e)) = s_{\mathcal{G}}(g(e))$, or
    $(h_{V,ext};[h_{1},h_{2}] + id_{n})(s_{1}(e)) = (h_{V,ext};[h_{1},h_{2}] + id_{n})(s_{2}(e))$.
    For the latter to hold when $s_{1}(e) \not = s_{2}(e)$ it must be the case that $h_{V,ext};[h_{1},h_{2}]$ identifies $v_{1} \in s_{1}(e)$ with $v_{2} \in s_{2}(e)$ and necessarily $v_{1}$ and $v_{2}$ are in the image of $c$.
    This can only happen if $v_1 \in i$ and $v_2 \in j$, \textit{i.e.} they belong to input and output interfaces respectively.
    However, this contradicts the fact that $v_2 \in s_2(e)$ as a vertex which is a source of some edge can not be in the output interface as per the (6) and (7) conditions of the boundary complement~\ref{def:boundary_new} the image of $j$ in $\mathcal{L}^{\bot}$ should consist of vertices of in-degree 0.
    Similarly for $t_{\mathcal{L}^{\bot}}$.

    % Because $g$ is a homomorphism, it must be the case that $g_{V}^{*}(s_{\mathcal{L^{\bot}}}(e)) = s_{\mathcal{G}}(g_{E}(e))$ and $s_{\mathcal{L^{\bot}}}(e) \subseteq (g_{V}^{-1})^{*}(s_{\mathcal{G}}(g_{E}(e)))$.
    % Since $g_{V} = h_{V,ext};[h_1,h_2] + id_{n}$ the inverse image of $g_{V}$ contains at most two elements: one with the pre-image in $i$ and another with the pre-image in $j$.
    % If $g_{V}$ identified more than two vertices it would mean that $[h_1,h_2]$ identified more than two vertices which would mean that two of them come both either from $i$ or $j$ which would imply that either $h_1$ or $h_2$ is not mono.
    % If the inverse image contains exactly one vertex then the $s_{\mathcal{L^{\bot}}}(e)$ is uniquely fixed, otherwise suppose it contains $\{v_1,v_2\}$ such that $v_1 \in i$ and $v_2 \in j$.
    % However, $v_1$ can not be in $s_{\mathcal{L^{\bot}}}(e)$ as per the (6) and (7) conditions of the boundary complement~\ref{def:boundary_new} the image of $j$ in $\mathcal{L}^{\bot}$ should consist of vertices of in-degree 0.

@@

    which implies $<_{2}(\iota_{V_{\mathcal{L}^{\bot}}}(v)) = <_{1}(\iota_{V_{\mathcal{L}^{\bot}}}(v))$ for all $v$ such that $[v) \not = \varnothing$ since $g_{E}$ \textcolor{red}{only identifies edges with no successors}.

@@

    Clearly there is a morphism from $\mathcal{L}^{\bot}$ to $\mathcal{G}'$, i.e. the morphism $g' = (g_{V},g_{E})$ that has the same action on edges and vertices as the morphism $g$ does, because $v'$ such that $[v') = \varnothing$ can be mapped to $g_{V}(v')$ such that $[g_{V}(v')) \not = \varnothing$.

@@

    The same reasoning applies when we consider $<_{1}(\iota_{E_{\mathcal{L}^{\bot}}}(e))$ and $<_{2}(\iota_{E_{\mathcal{L}^{\bot}}}(e))$.

    Finally, let's suppose there exist
    \[
        \mathcal{L}^{\bot}_{1} = (V_{\mathcal{L}^{\bot}}, E_{\mathcal{L}^{\bot}}, s,t, <, \consistency_{1})
    \quad
    \text{and}
    \quad
        \mathcal{L}^{\bot}_{2} = (V_{\mathcal{L}^{\bot}},V_{\mathcal{L}^{\bot}},s,t,<, \consistency_{2})
    \]
    Because $g$ is a homomorphism, we have
    \[
        [~g_{V};\iota_{V_{\mathcal{G}}}, g_{E};\iota_{E_{\mathcal{G}}}~]^{*}(\consistency_{\mathcal{L}^{\bot}_{1}}(\iota_{V_{\mathcal{L}^{\bot}}}(v))) \subseteq \consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v))
    \]
    \[
        [~g_{V};\iota_{V_{G}}, g_{E};\iota_{E_{\mathcal{G}}}]^{*}(\consistency_{\mathcal{L}^{\bot}_{2}}(\iota_{V_{\mathcal{L}^{\bot}}}(v))) \subseteq \consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v))
    \]
    \begin{itemize}
        \item If $\consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v)) = \varnothing$ then both left-hand sides should be $\varnothing$ and therefore the relations are equal.
        \item Suppose that $\consistency_{\mathcal{G}}(g_{V};\iota_{V_{\mathcal{G}}}(v)) \not = \varnothing$ which further means that $[g_{V};\iota_{V_{\mathcal{G}}}(v)) \not = \varnothing$.
              $\consistency_{1} \not = \consistency_{2}$ implies there exist $v, v' \in \mathcal{L}^{\bot}$ such that $v \consistency_{1} v'$ and $v \not \consistency_{2} v'$ for $[v) = [v') \not = \varnothing$.
              Then there are connected components $C_1, C_2$ in $\mathcal{L}_{2}^{\bot}$ such that for all $x \in C_1$ and $y \in C_2$ $x \not \consistency_{2} y$ and $v \in C_1$, $v' \in C_2$.
              Consider $g_2 = (g_{V},g_{E}) : \mathcal{L}_2^{\bot} \to \mathcal{G}$ and construct $\mathcal{G'}$ that is obtained from $\mathcal{G}$ by making $g(x) \not \consistency g(y)$ for $x \in C_1$ and $y \in C_2$.
              There is a morphism $g_2' = g_2 : \mathcal{L}_2^{\bot} \to \mathcal{G}'$. Like above, the only case when such morphism may not exist is when for $v' \not \consistency v$ $g_{V}(v) = g_{V}(v')$.
              This implies $v, v'$ have a pre-image in $i + j$ and by definition it must be $v' \consistency v$.
              Similarly, there is a morphism $m' : \mathcal{L} \to \mathcal{G}'$. The only case when it may not exist is when for $v_1, v_1' \in \mathcal{L}$ such that $v_1 \consistency v_1'$, $m_{V}(v_1) = g_{V}(v_2)$ and $m_{V}(v_1') = g_{V}(v_2')$ where $v_2 \in C_1$ and $v_2' \in C_2$.
              It implies $v_1, v_1'$ have pre-image in $i + j$ and by definition of a boundary-complement the image of $i + j$ should consist of top-level vertices exclusively.
              This results in the diagram below there is no morphism $u : G \to G'$, as the existence of such a morphism implies $u(g(x)) = u(g'(x))$, $u(g(y)) = u(g'(y))$ for $x \in C_1$ and $y \in C_2$ and $u(g(x)) \consistency u(g(y))$ and $u(g'(x)) \not \consistency u(g'(y))$.
              This contradicts the fact that $\mathcal{G}$ is a pushout.
              Hence, $\consistency_{1} = \consistency_{2}$ and $\mathcal{L}_{2}^{\bot} = \mathcal{L}_{1}^{\bot}$.
    \end{itemize}
    Ultimately, we have shown that $\mathcal{L}^{\bot}$ is unique up to isomorphism.
    The same reasoning applies when we consider edges instead of vertices.
\end{proof}

@@

\begin{lemma} Two e-classes $e_1$ and $e_2$ are merged if and only if there are rewrite rules that turn every term representable by $e_1$ into $e_2$.

\begin{lemma}
Every hierarchical edge can be decomposed as follows.

\[
\tikzfig{../figures/appendix/hyperedge_decomposition}
\]
\end{lemma}

Then, given e-classes $e_1$ and $e_2$ for sub-hypergraphs $C$ and $D$ respectively, we introduce e-nodes into the e-graph $A(e_1)$ and $B(e_2)$ that reside in e-class $e_3$.

### 2024-12-24 LICS/sections/e-hypergraphs.tex
makefile added

Note that here,  and in the rest of the paper,  we will restrict to dealing with signatures $\Sigma$  such that for all $c_1,c_2 \in \Sigma$, if arity (respectively, co-arity) of $c_1$ is 0, then co-arity (respectively, arity) of $c_2$ must be at least 1.
This is to ensure we do not have terms of type $0 \to 0$.
This is a restriction we make in order to avoid complications to the technical development of DPOI rewriting for e-hypergraphs given in the subsequent section,  and we explain it further there.

@@

\begin{definition}[E-hypergraphs]
\label{def:e-homo}    
An \emph{e-hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l,<,\consistency)$,  where $(V,E,s,t)$ is an unlabelled hypergraph;  $l :E \to \Sigma + 1$ is a labelling function modified to include an extra value $\bot$ in its codomain; $<$ is a strict partial order on $V + E$ called the \textit{child relation}.  
We define the following additional notation.  
For $x \in V + E$,  the set of \emph{parents} of $x$, is denoted as $[x) = \{x' ~|~ x' < x\; \}$.
% \[
% 	[x) = \{x' ~|~ x' <_c x\; \} \qquad (x] = \{x' ~|~  x <_{c} x'\}
% \]
We will denote the immediate parent $e$ of $x$ by writing $e <^{\mu} x$.
We call edges $e$ such that $l(e) = \bot$ \textit{hierarchical},  and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \emph{top-level}.
We require that the child relation satisfies the following conditions:
\begin{enumerate}
	\item each parent set contains exclusively hierarchical edges;
	\item Each $x$ has at most one immediate parent;
	\item if $v \in s(e)$ then $e' <^{\mu} e$ iff $e' <^{\mu} v$ and similarly if $v \in t(e)$.
\end{enumerate}
We additionally define an equivalence relation $\consistency_p$,  called the \textit{consistency relation},  on each set $\{x \in V + E ~|~ p <^\mu x\}$ of elements which share the same parent.  
We collect the union of all these equivalence relations into one relation $\consistency = \bigcup_{p} \consistency_p$,  which is defined on a subset of $V+E$. 
Note that this relation is symmetric and transitive if considered on $V + E$.
The consistency relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ and there exists $e'$ such that $e' <^{\mu} v$ and $e' <^{\mu} e$ then $v \consistency e$.
\label{def:consistency_properties}
\end{definition}
Intuitively,  the child relation equips the hypergraph with a hierarchical structure: certain edges,  labelled with $\bot$,  are meant to interpret the dashed e-boxes of  string diagrams.  
These edges can be designated as "parents" to sub-e-hypergraphs,  which interpret the various components of the e-boxes.  These components are distinguished by the consistency relation, which partitions the maximal sub-e-hypergraph below the hierarchical edge.  
Technically,  condition $(3)$ on the child relations says that $<^{\mu}$ respects connectivity of edges and vertices,  and similarly for the consistency relation. 
\begin{definition}[E-hypergraph homomorphism]
An \textit{e-hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
\begin{enumerate}
    \item $\phi$ is a hypergraph homomorphism
    \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
	    \[
        \text{if } e <_{\mathcal{F}}^{\mu} v \text{ then } \phi(e) <_{\mathcal{G}}^{\mu} \phi(v)
        \]
        and for $e_1, e_2 \in E_{\mathcal{F}}$
        \[
        \text{if } e_1 <_{\mathcal{F}}^{\mu} e_2 \text{ then } \phi(e_1) <_{\mathcal{G}}^{\mu} \phi(e_2)
        \]
        \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$. 
\end{enumerate}
% If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
\end{definition}
The conditions on e-hypergraph homomorphisms require to preserve immediate parents and the consistency relation.  
\begin{definition}[Category of e-hypergraphs]
The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
\end{definition}

@@

\begin{remark}
When we wish to consider $\phi(\mathcal{F}) \subseteq \mathcal{G}$ as an e-hypergraph in its own right,  we let the immediate parent to be undefined for all $x$ whose immediate parent in $\mathcal{G}$ is outside $\phi(\mathcal{F})$.
\end{remark}

When modelling enriched string diagrams using e-hypergraphs,  we require to keep track of more than simply the ultimate external inputs and outputs of the diagram: each internal e-box has its own inputs and outputs,  which we must also keep track of.  Thus,  we introduce an \textit{extended} cospan construction.  Since the $\catname{EHyp(\Sigma)}$ does not have pushouts in general,  we skip straight to considering a category of extended cospans of discrete e-hypergraphs (\textit{i.e.},  discrete hypergraphs considered as e-hypergraphs in the obvious way),  for which pushouts do exist.

>>>>

Note that here,  and in the rest of the paper,  we will restrict to dealing with signatures $\Sigma$  such that for all $c_1,c_2 \in \Sigma$, if arity (respectively, co-arity) of $c_1$ is 0, then co-arity (respectively, arity) of $c_2$ must be at least 1.
This is to ensure we do not have terms of type $0 \to 0$.
This is a restriction we make in order to avoid complications to the technical development of DPOI rewriting for e-hypergraphs given in the subsequent section,  and we explain it further there.
\textcolor{red}{The definitions and proofs without this extra assumption are given in the Appendix.}

@@

\begin{definition}[E-hypergraphs]
\label{def:e-homo}    
An \emph{e-hypergraph $\mathcal{G}$ over a signature $\Sigma$} is a tuple $(V,E,s,t,l,<,\consistency)$,  where $(V,E,s,t)$ is an unlabelled hypergraph;  $l :E \to \Sigma + 1$ is a labelling function modified to include an extra value $\bot$ in its codomain; $<$ is a strict partial order on $V + E$ called the \textit{child relation}, for which we introduce the following additional notation.
For $x \in V + E$,  the set of \emph{parents} of $x$, is denoted as $[x) = \{x' ~|~ x' < x\; \}$.
% \[
% 	[x) = \{x' ~|~ x' <_c x\; \} \qquad (x] = \{x' ~|~  x <_{c} x'\}
% \]
We will denote the immediate parent $e$ of $x$ by writing $e <^{\mu} x$.
We call edges $e$ such that $l(e) = \bot$ \textit{hierarchical},  and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \emph{top-level}.
We require that the child relation satisfies the following conditions:
\begin{enumerate}
	\item each parent set contains exclusively hierarchical edges;
	\item Each $x$ has at most one immediate parent;
	\item The maximal edges must be labelled;
	\item if $v \in s(e)$ then $e' <^{\mu} e$ iff $e' <^{\mu} v$ and similarly if $v \in t(e)$.
\end{enumerate}
Finally, $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p <^\mu x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p <^{\mu}(v)$ and $p <^{\mu}(e)$ then $v \consistency_{p} e$.
We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p <^{\mu} v\}$ and $E_{p} = \{ e ~ | ~ p <^{\mu} e\}$.
% We collect the union of all these equivalence relations into one relation $\consistency = \bigcup_{p} \consistency_p$,  which is defined on a subset of $V+E$. 
% Note that this relation is symmetric and transitive if considered on $V + E$.
% The consistency relation must additionally satisfy that if $v \in s(e)$ or $v \in t(e)$ and there exists $e'$ such that $e' <^{\mu} v$ and $e' <^{\mu} e$ then $v \consistency e$.
\label{def:consistency_properties}
\end{definition}
Intuitively,  the child relation equips a hypergraph with a hierarchical structure: certain edges,  labelled with $\bot$,  are meant to interpret the dashed e-boxes of string diagrams.  
These edges can be designated as "parents" to sub-e-hypergraphs,  which interpret the various components of the e-boxes.  
These components are distinguished by the consistency relation, which partitions the maximal sub-e-hypergraph below the hierarchical edge.
% Technically,  condition $(4)$ on the child relations says that $<^{\mu}$ respects connectivity of edges and vertices,  and similarly for the consistency relation. 
An example of an e-hypergraph can be seen below.
\[
\adjustbox{scale=0.75}{
\tikzfig{../figures/combinatorial_semantics/e-hypergraph-example-2}
}
\]

This e-hypergraph corresponds to the following tuple
\[
(V = (v_1, v_2, v_3, v_4, v_5, v_6\}, E = \{e_1, e_2, e_3\}, s, t, l, <, \consistency)
\]
\begin{align*}
s(e_1) = \{v_1\} & t(e_1) = \{v_2\} & l(e_1) = \bot \\
s(e_2) = \{v_3\} & t(e_2) = \{v_4\} & l(e_2) = e_2 \\
s(e_3) = \{v_5\} & t(e_3) = \{v_6\} & l(e_3) = e_3 \\
\end{align*}

and $[v_3) = [v_4) = [v_5) = [v_6) = [e_2) = [e_3) = \{e_1\}$ and $[v_1) = [v_2) = [e_1) = \varnothing$.
$v_3 \consistency v_4 \consistency e_2$ and $v_5 \consistency e_3 \consistency v_6$ and the relation is undefined for the rest vertices and edges.
Because $v_3 \not \consistency v_5$ and because the relation is closed under connectivity we delimit the connected components by a dashed vertical line.
The condition $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$, in particular, does not allow hierarchical edges with no delimiting lines.
$e_2$ and $e_3$ are maximal edges.
\begin{definition}[E-hypergraph homomorphism]
An \textit{e-hypergraph homomorphism} $\phi: \mathcal{F} \to \mathcal{G}$ of e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
\begin{enumerate}
    \item $\phi$ is a hypergraph homomorphism
    \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
	    \[
        \text{if } e <_{\mathcal{F}}^{\mu} v \text{ then } \phi(e) <_{\mathcal{G}}^{\mu} \phi(v)
        \]
        and for $e_1, e_2 \in E_{\mathcal{F}}$
        \[
        \text{if } e_1 <_{\mathcal{F}}^{\mu} e_2 \text{ then } \phi(e_1) <_{\mathcal{G}}^{\mu} \phi(e_2)
        \]
        \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$. 
\end{enumerate}
% If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
\end{definition}
The conditions on e-hypergraph homomorphisms require to preserve immediate parents and the consistency relation.  
\begin{definition}[Category of e-hypergraphs]
The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
\end{definition}

@@

% This remark is confusing
% \begin{remark}
% When we wish to consider $\phi(\mathcal{F}) \subseteq \mathcal{G}$ as an e-hypergraph in its own right,  we let the immediate parent to be undefined for all $x$ whose immediate parent in $\mathcal{G}$ is outside $\phi(\mathcal{F})$.
% \end{remark}

When modelling enriched string diagrams using e-hypergraphs,  we require keeping track of more than simply the ultimate external inputs and outputs of the diagram: each internal e-box has its own inputs and outputs,  which we must also keep track of.  Thus,  we introduce an \textit{extended} cospan construction.  Since the $\catname{EHyp(\Sigma)}$ does not have pushouts in general,  we skip straight to considering a category of extended cospans of discrete e-hypergraphs (\textit{i.e.},  discrete hypergraphs considered as e-hypergraphs in the obvious way),  for which pushouts do exist.

### 2024-12-24 LICS/sections/introduction.tex
makefile added

A recently proposed,  alternative solution is \textit{equality saturation} 
\cite{10.1145/1594834.1480915}: instead of maintaining a \textit{single},  putatively optimised program which is rewritten \textit{destructively} at each step, a \textit{set} of equivalent programs is maintained instead, whereas each rewrite step \textit{non-destructively} grows the set.  Upon reaching a fixed point (\textit{saturation}),  a \textit{globally} optimal program can then be extracted from this set.
A naive approach is clearly unfeasible, since the size of this set grows exponentially with the number of rewrites. 
However, \textit{equality graphs (e-graphs)} \cite{EggPaper} are a data structure that can represent this set compactly, thus making the set tractable in practical applications.

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations. 
We propose to address this issue via a categorical axiomatisation for e-graphs and their rewriting. 
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories. 
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, as encountered in a large body of research. 
The starting innovation of our approach is the semilattice-enrichment. 
The \textit{join} operator lifts the formalism so that it can now operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.

>>>>

A recently proposed,  alternative solution is \textit{equality saturation} 
\cite{10.1145/1594834.1480915}: instead of maintaining a \textit{single},  putatively optimised program which is rewritten \textit{destructively} at each step, a \textit{set} of equivalent programs is maintained instead, whereas each rewrite step \textit{non-destructively} grows the set.  Upon reaching a fixed point (\textit{saturation}),  a \textit{globally} optimal program can then be extracted from this set.
A na\"ive approach is clearly unfeasible, since the size of this set grows exponentially with the number of rewrites. 
However, \textit{equality graphs (e-graphs)} \cite{EggPaper} are a data structure that can represent this set compactly, thus making the set tractable in practical applications.

Although equality saturation is already a state-of-the-art algorithmic optimisation technique, there is an opportunity to better understand its mathematical foundations. 
We propose to address this issue via a categorical axiomatisation for e-graphs and their rewriting. 
Considering programs as represented by terms of an arbitrary algebraic theory, we demonstrate a correspondence between e-graphs and (string diagrams for) simple \textit{semilattice-enriched} Cartesian categories. 
The approach we take is guided by the established methodology of creating a correspondence between string diagrams and graph rewriting, as encountered in a large body of research. 
The starting innovation of our approach is the semilattice-enrichment. 
\textcolor{red}{The \textit{join} operator lifts the formalism} so that it can now operate with (non-empty) \textit{sets} of terms (string diagrams, morphisms); the rest of the paper is about working out the mathematical implications of introducing this enrichment to string diagrams and their concrete combinatorial representation.
